---
title: "Effects of severing on rhizome number and size analyses"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document contains the code used to run the analyses for Q4 in the manuscript, including code used to generate the values in Tables S8-S13.

Packages:

```{r pkgs, warning=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(glmmTMB)
library(lmerTest)
library(emmeans)
library(ggplot2)
library(ggeffects)
library(readxl)
library(MuMIn)
library(kableExtra)
library(gt)
library(webshot)
```

## Analyses for 1991

Read in data, select relevant variables and convert to numeric, filter out systems without leaf area measurements in 1990 and 1991.

```{r}

mayappleDatasetFull<-readxl::read_excel("mayapple_1990.xlsx", na=".")

mayappleData91 <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,Sex_90,Sever,Time, 
                LfFtot_91, LfBtot_91, Lftot_91,
                LftotF_92, LftotB_92, Lftot_92,
                brnoF_91, brnoB_91,
                brnoF_92, brnoB_92) %>% 
  mutate(Lftot_92=parse_double(Lftot_92), Lftot_91=parse_double(Lftot_91)) %>% 
  mutate(LftotB_92=parse_double(LftotB_92), LfBtot_91=parse_double(LfBtot_91) ) %>% 
  mutate(LftotF_92=parse_double(LftotF_92), LfFtot_91=parse_double(LfFtot_91) ) %>% 
  mutate(brnoF_92=parse_double(brnoF_92), brnoB_92=parse_double(brnoB_92) ) %>% 
  filter(!is.na(Lf_90)) %>% 
  filter(!is.na(Lftot_91))
```


### Contingency table 1991

Creating a variable that describes whether plants had zero systems, one descendant system or a front and back system in 1991. Use this to do a chi-square test.

```{r, warning=FALSE,message=FALSE}

mayappleData91 <- mayappleData91 %>% mutate(Status = case_when(LfFtot_91!=0 & LfBtot_91!=0 ~"Two",
                                             LfFtot_91==0 & LfBtot_91!=0 ~"One",
                                             LfFtot_91!=0 & LfBtot_91==0 ~"One",
                                             LfFtot_91==0 & LfBtot_91==0 ~"None")) %>% 
  mutate(Status = factor(Status, levels=c("None", "One", "Two")))

chisq.test(mayappleData91$Status, mayappleData91$Sever)
chi<-chisq.test(mayappleData91$Status, mayappleData91$Sever)
chi$observed
chi$expected
chi$residuals
```

### Leaf area 1991

Working with the status variable created for the chi-squared analysis, and filtering out the `None` records, because they have zero leaf area by definition. Create a variable for the average leaf area of a system, which is the leaf area of the system for one descendant systems, and the leaf area divided by two for front and back systems.

```{r}
mayappleData91 <- mayappleData91 %>% filter(Status != "None") %>% 
  mutate(Lf91s=ifelse(Status=="Two", Lftot_91/2, Lftot_91)) 
```

Run linear mixed effects model, check residuals.
Run anova and estimate marginal means comparisons.
```{r}
s91mod2<-lmer(Lf91s~Status*Sever+(1|COLONY), data = mayappleData91)
par(mfrow=c(1,2))
hist(resid(s91mod2))
plot(predict(s91mod2),resid(s91mod2)) ; abline(h=0)

anova(s91mod2)
emmeans(s91mod2, pairwise~Status)
emmeans(s91mod2, pairwise~Sever|Status)
emmeans(s91mod2, pairwise~Status|Sever)
```


## Analyses for 1992

Taking the 1991 dataset and further filtering out records without a leaf area measurement in 1992.

```{r}
mayappleData92 <- mayappleData91 %>% 
  filter(!is.na(Lftot_92))
```

### Contingency table 1992

Creating a variable that describes whether plants had zero systems, one descendant system or a front and back system in 1992. Use this to do a chi-square test.


```{r, warning=FALSE,message=FALSE}

mayappleData92 <- mayappleData92 %>% mutate(Status = case_when(LftotF_92!=0 & LftotB_92!=0 ~"Two",
                                             LftotF_92==0 & LftotB_92!=0 ~"One",
                                             LftotF_92!=0 & LftotB_92==0 ~"One",
                                             LftotF_92==0 & LftotB_92==0 ~"None")) %>% 
  mutate(Status = factor(Status, levels=c("None", "One", "Two")))

chisq.test(mayappleData92$Status, mayappleData92$Sever)
chi<-chisq.test(mayappleData92$Status, mayappleData92$Sever)
chi$observed
chi$expected
chi$residuals
```


### Leaf area 1992

Working with the status variable created for the chi-squared analysis, and filtering out the `None` records, because they have zero leaf area by definition. Create a variable for the average leaf area of a system, which is the leaf area of the system for one descendant systems, and the leaf area divided by two for front and back systems.

```{r}
mayappleData92 <- mayappleData92 %>% filter(Status != "None") %>% 
  mutate(Lf92s=ifelse(Status=="Two", Lftot_92/2, Lftot_92)) 
```

Run linear mixed effects model, check residuals.
Run anova and estimate marginal means comparisons.

```{r}
s92mod2<-lmer(Lf92s~Status*Sever+(1|COLONY), data = mayappleData92)
par(mfrow=c(1,2))
hist(resid(s91mod2))
plot(predict(s91mod2),resid(s91mod2)) ; abline(h=0)

anova(s92mod2)
emmeans(s92mod2, pairwise~Status)
emmeans(s92mod2, pairwise~Sever|Status)
emmeans(s92mod2, pairwise~Status|Sever)
```

