---
title: "Effects of severing on rhizome number and size analyses"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document contains the code used to run the analyses for Q4 in the manuscript, including code used to generate the values in Tables S8-S13.

Packages:

```{r pkgs, warning=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(glmmTMB)
library(lmerTest)
library(emmeans)
library(ggplot2)
library(ggeffects)
library(MuMIn)
library(kableExtra)
library(gt)
library(webshot)
```

## Analyses for year t+1

Read in data, select relevant variables and convert to numeric, filter out systems without leaf area measurements in year $t$ and year $t+1$

```{r}

mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData_t1 <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time, 
                LfFtot_t1, LfBtot_t1, Lftot_t1,
                LftotF_t2, LftotB_t2, Lftot_t2,
                brnoF_t1, brnoB_t1,
                brnoF_t2, brnoB_t2) %>% 
  mutate(Lftot_t2=parse_double(Lftot_t2), Lftot_t1=parse_double(Lftot_t1)) %>% 
  mutate(LftotB_t2=parse_double(LftotB_t2), LfBtot_t1=parse_double(LfBtot_t1) ) %>% 
  mutate(LftotF_t2=parse_double(LftotF_t2), LfFtot_t1=parse_double(LfFtot_t1) ) %>% 
  mutate(brnoF_t2=parse_double(brnoF_t2), brnoB_t2=parse_double(brnoB_t2) ) %>% 
  filter(!is.na(Lf_t)) %>% 
  filter(!is.na(Lftot_t1))
```


### Contingency table year t+1

Creating a variable that describes whether plants had zero systems, one descendant system or a front and back system in year $t+1$. Use this to do a chi-square test.

```{r, warning=FALSE,message=FALSE}

mayappleData_t1 <- mayappleData_t1 %>% mutate(Status = case_when(LfFtot_t1!=0 & LfBtot_t1!=0 ~"Two",
                                             LfFtot_t1==0 & LfBtot_t1!=0 ~"One",
                                             LfFtot_t1!=0 & LfBtot_t1==0 ~"One",
                                             LfFtot_t1==0 & LfBtot_t1==0 ~"None")) %>% 
  mutate(Status = factor(Status, levels=c("None", "One", "Two")))

chisq.test(mayappleData_t1$Status, mayappleData_t1$Sever)
chi<-chisq.test(mayappleData_t1$Status, mayappleData_t1$Sever)
chi$observed
chi$expected
chi$residuals
```

### Leaf area year t+1

Working with the status variable created for the chi-squared analysis, and filtering out the `None` records, because they have zero leaf area by definition. Create a variable for the average leaf area of a system, which is the leaf area of the system for one descendant systems, and the leaf area divided by two for front and back systems.

```{r}
mayappleData_t1 <- mayappleData_t1 %>% filter(Status != "None") %>% 
  mutate(Lf_t1s=ifelse(Status=="Two", Lftot_t1/2, Lftot_t1)) 
```

Run linear mixed effects model, check residuals.
Run anova and estimate marginal means comparisons.
```{r}
s_t1mod2<-lmer(Lf_t1s~Status*Sever+(1|COLONY), data = mayappleData_t1)
par(mfrow=c(1,2))
hist(resid(s_t1mod2))
plot(predict(s_t1mod2),resid(s_t1mod2)) ; abline(h=0)

anova(s_t1mod2)
emmeans(s_t1mod2, pairwise~Status)
emmeans(s_t1mod2, pairwise~Sever|Status)
emmeans(s_t1mod2, pairwise~Status|Sever)
```


## Analyses for year t+2

Taking the year $t+1$ dataset and further filtering out records without a leaf area measurement in year $t+2$.

```{r}
mayappleData_t2 <- mayappleData_t1 %>% 
  filter(!is.na(Lftot_t2))
```

### Contingency table year t+2

Creating a variable that describes whether plants had zero systems, one descendant system or a front and back system in year $t+2$. Use this to do a chi-square test.


```{r, warning=FALSE,message=FALSE}

mayappleData_t2 <- mayappleData_t2 %>% mutate(Status = case_when(LftotF_t2!=0 & LftotB_t2!=0 ~"Two",
                                             LftotF_t2==0 & LftotB_t2!=0 ~"One",
                                             LftotF_t2!=0 & LftotB_t2==0 ~"One",
                                             LftotF_t2==0 & LftotB_t2==0 ~"None")) %>% 
  mutate(Status = factor(Status, levels=c("None", "One", "Two")))

chisq.test(mayappleData_t2$Status, mayappleData_t2$Sever)
chi<-chisq.test(mayappleData_t2$Status, mayappleData_t2$Sever)
chi$observed
chi$expected
chi$residuals
```


### Leaf area year t+2

Working with the status variable created for the chi-squared analysis, and filtering out the `None` records, because they have zero leaf area by definition. Create a variable for the average leaf area of a system, which is the leaf area of the system for one descendant systems, and the leaf area divided by two for front and back systems.

```{r}
mayappleData_t2 <- mayappleData_t2 %>% filter(Status != "None") %>% 
  mutate(Lf_t2s=ifelse(Status=="Two", Lftot_t2/2, Lftot_t2)) 
```

Run linear mixed effects model, check residuals.
Run anova and estimate marginal means comparisons.

```{r}
s_t2mod2<-lmer(Lf_t2s~Status*Sever+(1|COLONY), data = mayappleData_t2)
par(mfrow=c(1,2))
hist(resid(s_t2mod2))
plot(predict(s_t2mod2),resid(s_t2mod2)) ; abline(h=0)

anova(s_t2mod2)
emmeans(s_t2mod2, pairwise~Status)
emmeans(s_t2mod2, pairwise~Sever|Status)
emmeans(s_t2mod2, pairwise~Status|Sever)
```

