---
title: "Model selection analyses"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

This document contains:

- Code used to run the analyses that address Q1 and Q2 in the manuscript
- Code used to generate supplemental tables S1-S7

Packages:

```{r pkgs, warning=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(glmmTMB)
library(lmerTest)
library(emmeans)
library(ggplot2)
library(ggeffects)
library(MuMIn)
library(kableExtra)
library(gt)
library(webshot)
```

## Questions and Analyses

Model selection was performed on seven response variables that were measured between year $t$ and $t+2$ 

*Q1: Do sexual status and the position and timing of severing affect front and/or back shoot production and performance in the year of severing and in the following years?*

To address this question, we performed model selection on five response variables: 
1. leaf area in year $t$ (the year of severing)
2. the production of a front shoot(s) in year $t+1$ 
3. the total leaf area of the front system in year $t+1$ (for systems with shoots in year $t+1$) 
4. the production of a back shoot(s) in year $t+1$ 
5. the total leaf area of the back system in year $t+1$ (for systems with shoots in year $t+1$) 

*Q2: In the two years following the experimental severing, is plant performance still affected by the location and time of severing or is the previous year’s shoot size a better predictor of performance?*

To address this question, we used the model selection results from the four year $t+1$ response variables (see #2-#5 below) and two response variables that were measured in year $t+2$: total leaf area of the front portion of the system in year $t+2$ (#6), and total leaf area of the back portion of the system in year $t+2$ (#7). 

### 1

Leaf area in year $t$

Load data, make variable of interest numeric, filter out systems with zero leaf area:
```{r load year_t, warning=FALSE, message=FALSE}
may<-read_csv("mayapple.csv",na=".")

may$Lf_t<-as.numeric(as.character(may$Lf_t))

may<-may %>% 
  filter(!is.na(Lf_t))
```

Model specification:

```{r, warning=FALSE, message=FALSE}
model<-lmer(Lf_t~Sex_t*Sever*Time+(1|COLONY), data=may, na.action = "na.fail", REML=FALSE)
```

Model selection and creation of Table S1:

```{r, warning=FALSE, message=FALSE}

dr<-dredge(model)

tabS1 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,  Sever, Sex_t, Time,  `Sever:Sex_t`, `Sever:Time`, `Sex_t:Time`, `Sever:Sex_t:Time`)) %>% 
  tab_spanner(label = "Fixed parameter", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```

Specification of the best model:

```{r}
model.bm<-lmer(Lf_t~Sex_t*Time+Sever*Time+(1|COLONY), data=may, na.action = "na.fail", REML=FALSE)
```

Estimated marginal means for the main effects and interactions in the best model:

```{r}
#Sex_t: 
emmeans(model.bm, "Sex_t", type="response")

#Sever:
emmeans(model.bm, pairwise~Sever, type="response")

#Time:
emmeans(model.bm, pairwise~Time, type="response")

#Sex_t x Time:
emmeans(model.bm, pairwise~Time|Sex_t, type="response")

#Sever x Time:
emmeans(model.bm, pairwise~Time|Sever, type="response")

```

### 2

The production of a front shoot(s) in year $t+1$

Read in data, select variables from the front of the system. Also filtering out rows that don't have a leaf area record in year $t$. Create binary variable for production of one or more front shoots in year $t+1$. 

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,brnoF_t1,Sex_t,Sever,Time,LfFtot_t1) %>% 
  filter(!is.na(Lf_t))

mayappleData<- mayappleData %>% 
  dplyr::mutate(deadAliveBinary=ifelse(brnoF_t1>0,1,0))

```

Specify global model, perform model selection, generate Table S2.

```{r, warning=FALSE, message=FALSE}
modt1sur0l<-glmmTMB(deadAliveBinary~ Sex_t*Sever + Sex_t*Time + Sever*Time + Lf_t + (1|COLONY), data = mayappleData,
                family="binomial")

dr<-dredge(modt1sur0l)

tabS2 <- head(dr, n=10) %>% gt() %>% cols_hide(columns=vars(`disp((Int))`)) %>% 
  cols_label(`cond((Int))`="(Intercept)", `cond(Lf_t)`="Lf_t", `cond(Sever)`="Sever", `cond(Sex_t)`="Sex_t", `cond(Time)`="Time", `cond(Sever:Sex_t)`="Sever:Sex_t", `cond(Sever:Time)`="Sever:Time", `cond(Sex_t:Time)`= "Sex_t:Time") %>% 
  fmt_number(columns = vars(`cond((Int))`, logLik, AICc, delta, weight), decimals = 2) %>% 
  fmt_number(columns = vars(`cond(Lf_t)`), decimals = 4) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`cond((Int))`,  `cond(Lf_t)`, `cond(Sever)`, `cond(Sex_t)`, `cond(Time)`, `cond(Sever:Sex_t)`, `cond(Sever:Time)`, `cond(Sex_t:Time)`)) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`cond((Int))`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 2)) 

```

Specification of the best model:
```{r}
sur_t1f_bm<-glmmTMB(deadAliveBinary~ Sever+Time +  (1|COLONY), data = mayappleData,
                family="binomial")
```

Estimated marginal means for the main effects:

```{r}
#Sever:
emmeans(sur_t1f_bm, pairwise~Sever, type="response" )

#Time:
emmeans(sur_t1f_bm, pairwise~Time, type="response" )
```

### 3

The total leaf area of the front system in year $t+1$ (for systems with shoots in year $t+1$) 

Read in data, select relevant variables. Convert response to numeric, remove NA values and select systems with non-zero back leaf area in year $t+1$.

```{r}

mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,brnoF_t1,Sex_t,Sever,Time,LfFtot_t1) %>% 
  filter(!is.na(Lf_t))

mayappleData$LfFtot_t1<-as.numeric(mayappleData$LfFtot_t1)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfFtot_t1)) %>% 
  filter(LfFtot_t1>0)
```

Specify global model, perform model selection, generate Table S3:

```{r, warning=FALSE, message=FALSE}

modt1laf<-lmer(LfFtot_t1~Sex_t*Sever*Time+Lf_t +(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

dr<-dredge(modt1laf)

tabS3 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, Lf_t,  logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,   Lf_t,  Sever, Sex_t, Time,  `Sever:Sex_t`, `Sever:Time`, `Sex_t:Time`, `Sever:Sex_t:Time`)) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 4)) 
```

Specify best model: 

```{r}
modt1laf.bm<-lmer(LfFtot_t1~Sever+Lf_t +(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)
```

Estimate marginal means/partial regression coefficients for the main effects:

```{r}
#Sever:
emmeans(modt1laf.bm, pairwise~Sever, type="response")
#Leaf area in year t (see estimate for Lf_t in summary output)
summary(modt1laf.bm)
```


### 4

The production of a back shoot(s) in year $t+1$

Read in data, select relevant variables, filter out records with zero leaf area in year $t$. Create binary variable of back shoot status in year $t+1$.

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time,brnoB_t1,LfBtot_t1,Lftot_t1) %>% 
  filter(!is.na(Lf_t))

mayappleData<- mayappleData %>% 
  dplyr::mutate(backShootBranchBinary=ifelse(brnoB_t1>0,1,0)) %>% 
  filter(!is.na(backShootBranchBinary))
```

Specify global model, perform model selection, generate Table S4:

```{r}
b_modt1sur<-glmmTMB(backShootBranchBinary~ Sex_t*Sever + Sex_t*Time + Lf_t + (1|COLONY), data = mayappleData,
                family="binomial")

dr<-dredge(b_modt1sur)

tabS4 <- head(dr, n=10) %>% gt() %>% cols_hide(columns=vars(`disp((Int))`)) %>% 
  cols_label(`cond((Int))`="(Intercept)", `cond(Lf_t)`="Lf_t", `cond(Sever)`="Sever", `cond(Sex_t)`="Sex_t", `cond(Time)`="Time", `cond(Sever:Sex_t)`="Sever:Sex_t",  `cond(Sex_t:Time)`= "Sex_t:Time") %>% 
  fmt_number(columns = vars(`cond((Int))`, logLik, AICc, delta, weight), decimals = 2) %>% 
  fmt_number(columns = vars(`cond(Lf_t)`), decimals = 4) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`cond((Int))`,  `cond(Lf_t)`, `cond(Sever)`, `cond(Sex_t)`, `cond(Time)`, `cond(Sever:Sex_t)`,  `cond(Sex_t:Time)`)) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`cond((Int))`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```

Specify best model: 

```{r}
sur_t1b_bm<-glmmTMB(backShootBranchBinary~ Sever+Time + Lf_t +(1|COLONY), data = mayappleData,
                family="binomial")
```

Estimate marginal means/partial regression coefficients for the main effects:

```{r}
#Sever
emmeans(sur_t1b_bm, pairwise~Sever, type="response" )
#Time
emmeans(sur_t1b_bm, pairwise~Time, type="response" )
#Leaf area in year t (see estimate for Lf_t in summary output)
summary(sur_t1b_bm)
```

### 5

The total leaf area of the back system in year $t+1$ (for systems with shoots in year $t+1$)

Read in data, select relevant variables. Convert response to numeric, remove NA values and select systems with non-zero back leaf area in year $t+1$.
```{r}
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time,brnoB_t1,LfBtot_t1,Lftot_t1) %>% 
  filter(!is.na(Lf_t))

mayappleData$LfBtot_t1<-as.numeric(mayappleData$LfBtot_t1)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfBtot_t1)) %>% 
  filter(LfBtot_t1>0)

```

Specify global model, perform model selection, generate Table S5:

```{r, warning=FALSE, message=FALSE}

b_modt1la.0l<-lmer(LfBtot_t1~Sex_t*Time+Sever +Lf_t+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

dr<-dredge(b_modt1la.0l)

tabS5 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, Lf_t, logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,   Lf_t,  Sever, Sex_t, Time,    `Sex_t:Time` )) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```

Specify best model:

```{r}
b_modt1la.bm<-lmer(LfBtot_t1~Sex_t+Sever+Time +Lf_t+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)
```

Estimate marginal means/partial regression coefficients for the main effects:

```{r}
#Sex_t:
emmeans(b_modt1la.bm, pairwise~Sex_t, type="response")
#Time:
emmeans(b_modt1la.bm, pairwise~Time, type="response")
#Sever:
emmeans(b_modt1la.bm, pairwise~Sever, type="response")
#Leaf area in year t (see estimate for Lf_t in summary output)
summary(b_modt1la.bm)
```

### 6

The total leaf area of the front system in year $t+2$ (for systems with shoots in year $t+2$). This analysis uses all systems that have a non-NA value for total leaf area of the front and back in year $t+1$ and in year $t+2$, excluding those systems for which year $t$ leaf area was NA. *Furthermore*, it restricts the analysis to those systems that have a non-zero leaf area in year $t+2$

Read in data, select relevant variables. Filter out the records without leaf area measurements in year $t$ and/or year $t+1$ and/or year $t+2$ (front).

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time,LftotF_t2, Lftot_t1, LfBtot_t1, LfFtot_t1) %>% 
  mutate(LftotF_t2=parse_double(LftotF_t2), Lftot_t1=parse_double(Lftot_t1), LfBtot_t1=parse_double(LfBtot_t1), LfFtot_t1=parse_double(LfFtot_t1) ) %>% 
  filter(!is.na(Lf_t)) %>% 
  filter(!is.na(Lftot_t1)) %>% 
  filter(!is.na(LftotF_t2)) %>% 
  filter(LftotF_t2>0)
```

Specify global model, perform model selection, generate Table S6:

```{r, warning=FALSE, message=FALSE}
d_modt2la.0l<-lmer(LftotF_t2~Sex_t*Sever*Time +LfFtot_t1+Lf_t+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

dr<-dredge(d_modt2la.0l)

tabS6 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, Lf_t, LfFtot_t1, logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,   Lf_t, LfFtot_t1, Sever, Sex_t, Time,  `Sever:Sex_t`, `Sever:Time`, `Sex_t:Time`, `Sever:Sex_t:Time`)) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```

Specify best model:

```{r}
d_modt2la.bm<-lmer(LftotF_t2~+LfFtot_t1+Lf_t+(1|COLONY), data = mayappleData, na.action ="na.fail")
```

Estimate partial regression coefficients for the main effects:

```{r}
#Partial regression coefficients for Lf_t and LfFtot_t1 can be found in the model summary table:
summary(d_modt2la.bm)
```


### 7

The total leaf area of the back system in year $t+2$ (for systems with shoots in year $t+2$). This analysis uses all systems that have a non-NA value for total leaf area of the front and back in year $t+1$ and in year $t+2$, excluding those systems for which year $t$ leaf area was NA. *Furthermore*, this analysis excludes any systems with a zero leaf area in year $t+2$.

Read in data, select relevant variables. Filter out the records without leaf area measurements in year $t$ and/or year $t+1$ and/or year $t+2$ (back).

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time,LftotB_t2, Lftot_t1, LfBtot_t1, LfFtot_t1) %>% 
  mutate(LftotB_t2=parse_double(LftotB_t2), Lftot_t1=parse_double(Lftot_t1), LfBtot_t1=parse_double(LfBtot_t1), LfFtot_t1=parse_double(LfFtot_t1) ) %>% 
  filter(!is.na(Lf_t)) %>% 
  filter(!is.na(Lftot_t1)) %>% 
  filter(!is.na(LftotB_t2)) %>% 
  filter(LftotB_t2>0)
```

Specify global model, perform model selection, generate Table S7:

```{r, warning=FALSE, message=FALSE}
d_modt2la.0l<-lmer(LftotB_t2~Sex_t*Sever*Time +LfBtot_t1+Lf_t+(1|COLONY), data = mayappleData, na.action ="na.fail",REML=FALSE)

dr<-dredge(d_modt2la.0l)

tabS7 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, Lf_t, LfBtot_t1, logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,   Lf_t, LfBtot_t1, Sever, Sex_t, Time,  `Sever:Sex_t`, `Sever:Time`, `Sex_t:Time`, `Sever:Sex_t:Time`)) %>% 
  tab_spanner(label = "Fixed parameter", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```


Specify best model:

```{r}
d_modt2la.bm<-lmer(LftotB_t2~Sever +Lf_t+LfBtot_t1+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)
```

Estimate marginal means/partial regression coefficients for the main effects:

```{r}
#Sever:
emmeans(d_modt2la.bm, pairwise~Sever, type="response")
#Partial regression coefficient for Lf_t and LfBtot_t1 can be found in the model summary table:
summary(d_modt2la.bm)
```


