---
title: "Model selection analyses"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

This document contains:

- Code used to run the analyses that address Q1, Q2, and Q3 in the manuscript
- Code used to generate supplemental tables S1-S7

Packages:

```{r pkgs, warning=FALSE, message=FALSE, include=FALSE}
library(tidyverse)
library(glmmTMB)
library(lmerTest)
library(emmeans)
library(ggplot2)
library(ggeffects)
library(MuMIn)
library(kableExtra)
library(gt)
library(webshot)
```

## Questions and Analyses

Model selection was performed on seven response variables that were measured between 1990 and 1992. 

*Q1: Do the position and timing of severing affect shoot performance in the year of severing and in the following year?*

To address this question, we performed model selection on five response variables: 
1. leaf area in 1990 (the year of severing)
2. the production of a front shoot(s) in 1991 
3. the total leaf area of the front system in 1991 (for systems with shoots in 1991) 
4. the production of a back shoot(s) in 1991 
5. the total leaf area of the back system in 1991 (for systems with shoots in 1991) 

*Q2: Do the position and timing of severing affect the ability of the back system to recover from damage by producing a back shoot?*

To address this question, we used the model selection results from the analysis of back shoot production in 1991. (See #4 below)

*Q3: In the years following the experimental severing, are there persistent effects of severing on plant performance or are these mediated through treatment effects on past years’ performance?*

To address this question, we used the model selection results from the four 1991 response variables (see #2-#5 below) and two response variables that were measured in 1992: total leaf area of the front portion of the system in 1992 (#6), and total leaf area of the back portion of the sytem in 1992 (#7). 

### 1

Leaf area in 1990

Load data, make variable of interest numeric, filter out systems with zero leaf area:
```{r load 1990, warning=FALSE, message=FALSE}
may<-read_csv("mayapple_1990.csv",na=".")

may$Lf_90<-as.numeric(as.character(may$Lf_90))

may<-may %>% 
  filter(!is.na(Lf_90))
```

Model specification:

```{r, warning=FALSE, message=FALSE}
model<-lmer(Lf_90~Sex_90*Sever*Time+(1|COLONY), data=may, na.action = "na.fail", REML=FALSE)
```

Model selection and creation of Table S1:

```{r, warning=FALSE, message=FALSE}

dr<-dredge(model)

tabS1 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,  Sever, Sex_90, Time,  `Sever:Sex_90`, `Sever:Time`, `Sex_90:Time`, `Sever:Sex_90:Time`)) %>% 
  tab_spanner(label = "Fixed parameter", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```

Specification of the best model:

```{r}
model.bm<-lmer(Lf_90~Sex_90*Time+Sever*Time+(1|COLONY), data=may, na.action = "na.fail", REML=FALSE)
```

Estimated marginal means for the main effects and interactions in the best model:

```{r}
#Sex_90: 
emmeans(model.bm, "Sex_90", type="response")

#Sever:
emmeans(model.bm, pairwise~Sever, type="response")

#Time:
emmeans(model.bm, pairwise~Time, type="response")

#Sex_90 x Time:
emmeans(model.bm, pairwise~Time|Sex_90, type="response")

#Sever x Time:
emmeans(model.bm, pairwise~Time|Sever, type="response")

```

### 2

The production of a front shoot(s) in 1991

Read in data, select variables from the front of the system. Also filtering out rows that don't have a leaf area record in 1990. Create binary variable for production of one or more front shoots in 1991. 

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple_1990.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,brnoF_91,Sex_90,Sever,Time,LfFtot_91) %>% 
  filter(!is.na(Lf_90))

mayappleData<- mayappleData %>% 
  dplyr::mutate(deadAliveBinary=ifelse(brnoF_91>0,1,0))

```

Specify global model, perform model selection, generate Table S2.

```{r, warning=FALSE, message=FALSE}
mod91sur0l<-glmmTMB(deadAliveBinary~ Sex_90*Sever + Sex_90*Time + Sever*Time + Lf_90 + (1|COLONY), data = mayappleData,
                family="binomial")

dr<-dredge(mod91sur0l)

tabS2 <- head(dr, n=10) %>% gt() %>% cols_hide(columns=vars(`disp((Int))`)) %>% 
  cols_label(`cond((Int))`="(Intercept)", `cond(Lf_90)`="Lf_90", `cond(Sever)`="Sever", `cond(Sex_90)`="Sex_90", `cond(Time)`="Time", `cond(Sever:Sex_90)`="Sever:Sex_90", `cond(Sever:Time)`="Sever:Time", `cond(Sex_90:Time)`= "Sex_90:Time") %>% 
  fmt_number(columns = vars(`cond((Int))`, logLik, AICc, delta, weight), decimals = 2) %>% 
  fmt_number(columns = vars(`cond(Lf_90)`), decimals = 4) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`cond((Int))`,  `cond(Lf_90)`, `cond(Sever)`, `cond(Sex_90)`, `cond(Time)`, `cond(Sever:Sex_90)`, `cond(Sever:Time)`, `cond(Sex_90:Time)`)) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`cond((Int))`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 2)) 

```

Specification of the best model:
```{r}
sur_91f_bm<-glmmTMB(deadAliveBinary~ Sever+Time +  (1|COLONY), data = mayappleData,
                family="binomial")
```

Estimated marginal means for the main effects:

```{r}
#Sever:
emmeans(sur_91f_bm, pairwise~Sever, type="response" )

#Time:
emmeans(sur_91f_bm, pairwise~Time, type="response" )
```

### 3

The total leaf area of the front system in 1991 (for systems with shoots in 1991) 

Read in data, select relevant variables. Convert response to numeric, remove NA values and select systems with non-zero back leaf area in 1991.

```{r}

mayappleDatasetFull<-read_csv("mayapple_1990.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,brnoF_91,Sex_90,Sever,Time,LfFtot_91) %>% 
  filter(!is.na(Lf_90))

mayappleData$LfFtot_91<-as.numeric(mayappleData$LfFtot_91)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfFtot_91)) %>% 
  filter(LfFtot_91>0)
```

Specify global model, perform model selection, generate Table S3:

```{r, warning=FALSE, message=FALSE}

mod91laf<-lmer(LfFtot_91~Sex_90*Sever*Time+Lf_90 +(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

dr<-dredge(mod91laf)

tabS3 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, Lf_90,  logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,   Lf_90,  Sever, Sex_90, Time,  `Sever:Sex_90`, `Sever:Time`, `Sex_90:Time`, `Sever:Sex_90:Time`)) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 4)) 
```

Specify best model: 

```{r}
mod91laf.bm<-lmer(LfFtot_91~Sever+Lf_90 +(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)
```

Estimate marginal means/partial regression coefficients for the main effects:

```{r}
#Sever:
emmeans(mod91laf.bm, pairwise~Sever, type="response")
#Leaf area in 1990 (see estimate for Lf_90 in summary output)
summary(mod91laf.bm)
```


### 4

The production of a back shoot(s) in 1991

Read in data, select relevant variables, filter out records with zero leaf area in 1990. Create binary variable of back shoot status in 1991.

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple_1990.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,Sex_90,Sever,Time,brnoB_91,LfBtot_91,Lftot_91) %>% 
  filter(!is.na(Lf_90))

mayappleData<- mayappleData %>% 
  dplyr::mutate(backShootBranchBinary=ifelse(brnoB_91>0,1,0)) %>% 
  filter(!is.na(backShootBranchBinary))
```

Specify global model, perform model selection, generate Table S4:

```{r}
b_mod91sur<-glmmTMB(backShootBranchBinary~ Sex_90*Sever + Sex_90*Time + Lf_90 + (1|COLONY), data = mayappleData,
                family="binomial")

dr<-dredge(b_mod91sur)

tabS4 <- head(dr, n=10) %>% gt() %>% cols_hide(columns=vars(`disp((Int))`)) %>% 
  cols_label(`cond((Int))`="(Intercept)", `cond(Lf_90)`="Lf_90", `cond(Sever)`="Sever", `cond(Sex_90)`="Sex_90", `cond(Time)`="Time", `cond(Sever:Sex_90)`="Sever:Sex_90",  `cond(Sex_90:Time)`= "Sex_90:Time") %>% 
  fmt_number(columns = vars(`cond((Int))`, logLik, AICc, delta, weight), decimals = 2) %>% 
  fmt_number(columns = vars(`cond(Lf_90)`), decimals = 4) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`cond((Int))`,  `cond(Lf_90)`, `cond(Sever)`, `cond(Sex_90)`, `cond(Time)`, `cond(Sever:Sex_90)`,  `cond(Sex_90:Time)`)) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`cond((Int))`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```

Specify best model: 

```{r}
sur_91b_bm<-glmmTMB(backShootBranchBinary~ Sever+Time + Lf_90 +(1|COLONY), data = mayappleData,
                family="binomial")
```

Estimate marginal means/partial regression coefficients for the main effects:

```{r}
#Sever
emmeans(sur_91b_bm, pairwise~Sever, type="response" )
#Time
emmeans(sur_91b_bm, pairwise~Time, type="response" )
#Leaf area in 1990 (see estimate for Lf_90 in summary output)
summary(sur_91b_bm)
```

### 5

The total leaf area of the back system in 1991 (for systems with shoots in 1991)

Read in data, select relevant variables. Convert response to numeric, remove NA values and select systems with non-zero back leaf area in 1991.
```{r}
mayappleDatasetFull<-read_csv("mayapple_1990.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,Sex_90,Sever,Time,brnoB_91,LfBtot_91,Lftot_91) %>% 
  filter(!is.na(Lf_90))

mayappleData$LfBtot_91<-as.numeric(mayappleData$LfBtot_91)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfBtot_91)) %>% 
  filter(LfBtot_91>0)

```

Specify global model, perform model selection, generate Table S5:

```{r, warning=FALSE, message=FALSE}

b_mod91la.0l<-lmer(LfBtot_91~Sex_90*Time+Sever +Lf_90+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

dr<-dredge(b_mod91la.0l)

tabS5 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, Lf_90, logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,   Lf_90,  Sever, Sex_90, Time,    `Sex_90:Time` )) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`(Intercept)`,Lf_90)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```

Specify best model:

```{r}
b_mod91la.bm<-lmer(LfBtot_91~Sex_90+Sever+Time +Lf_90+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)
```

Estimate marginal means/partial regression coefficients for the main effects:

```{r}
#Sex_90:
emmeans(b_mod91la.bm, pairwise~Sex_90, type="response")
#Time:
emmeans(b_mod91la.bm, pairwise~Time, type="response")
#Sever:
emmeans(b_mod91la.bm, pairwise~Sever, type="response")
#Leaf area in 1990 (see estimate for Lf_90 in summary output)
summary(b_mod91la.bm)
```

### 6

The total leaf area of the front system in 1992 (for systems with shoots in 1992). This analysis uses all systems that have a non-NA value for total leaf area of the front and back in 1991 and in 1992, excluding those systems for which 1990 leaf area was NA. *Furthermore*, it restricts the analysis to those systems that have a non-zero leaf area in 1992.

Read in data, select relevant variables. Filter out the records without leaf area measurements in 1990 and/or 1991 and/or 1992 (front).

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple_1990.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,Sex_90,Sever,Time,LftotF_92, Lftot_91, LfBtot_91, LfFtot_91) %>% 
  mutate(LftotF_92=parse_double(LftotF_92), Lftot_91=parse_double(Lftot_91), LfBtot_91=parse_double(LfBtot_91), LfFtot_91=parse_double(LfFtot_91) ) %>% 
  filter(!is.na(Lf_90)) %>% 
  filter(!is.na(Lftot_91)) %>% 
  filter(!is.na(LftotF_92)) %>% 
  filter(LftotF_92>0)
```

Specify global model, perform model selection, generate Table S6:

```{r, warning=FALSE, message=FALSE}
d_mod92la.0l<-lmer(LftotF_92~Sex_90*Sever*Time +LfFtot_91+Lf_90+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

dr<-dredge(d_mod92la.0l)

tabS6 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, Lf_90, LfFtot_91, logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,   Lf_90, LfFtot_91, Sever, Sex_90, Time,  `Sever:Sex_90`, `Sever:Time`, `Sex_90:Time`, `Sever:Sex_90:Time`)) %>% 
  tab_spanner(label = "Fixed parameters", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```

Specify best model:

```{r}
d_mod92la.bm<-lmer(LftotF_92~+LfFtot_91+Lf_90+(1|COLONY), data = mayappleData, na.action ="na.fail")
```

Estimate partial regression coefficients for the main effects:

```{r}
#Partial regression coefficients for Lf_90 and LfFtot_91 can be found in the model summary table:
summary(d_mod92la.bm)
```


### 7

The total leaf area of the back system in 1992 (for systems with shoots in 1992). This analysis uses all systems that have a non-NA value for total leaf area of the front and back in 1991 and in 1992, excluding those systems for which 1990 leaf area was NA. *Furthermore*, this analysis excludes any systems with a zero leaf area in 1992.

Read in data, select relevant variables. Filter out the records without leaf area measurements in 1990 and/or 1991 and/or 1992 (back).

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple_1990.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,Sex_90,Sever,Time,LftotB_92, Lftot_91, LfBtot_91, LfFtot_91) %>% 
  mutate(LftotB_92=parse_double(LftotB_92), Lftot_91=parse_double(Lftot_91), LfBtot_91=parse_double(LfBtot_91), LfFtot_91=parse_double(LfFtot_91) ) %>% 
  filter(!is.na(Lf_90)) %>% 
  filter(!is.na(Lftot_91)) %>% 
  filter(!is.na(LftotB_92)) %>% 
  filter(LftotB_92>0)
```

Specify global model, perform model selection, generate Table S7:

```{r, warning=FALSE, message=FALSE}
d_mod92la.0l<-lmer(LftotB_92~Sex_90*Sever*Time +LfBtot_91+Lf_90+(1|COLONY), data = mayappleData, na.action ="na.fail",REML=FALSE)

dr<-dredge(d_mod92la.0l)

tabS7 <- head(dr, n=10) %>% gt() %>%
  fmt_number(columns = vars(`(Intercept)`, Lf_90, LfBtot_91, logLik, AICc, delta, weight), decimals = 2) %>% 
  tab_spanner(label = "Fit information", columns = vars(df, logLik, AICc, delta, weight)) %>% 
  tab_spanner(label = "Non-fixed parameters", columns = vars(`(Intercept)`,   Lf_90, LfBtot_91, Sever, Sex_90, Time,  `Sever:Sex_90`, `Sever:Time`, `Sex_90:Time`, `Sever:Sex_90:Time`)) %>% 
  tab_spanner(label = "Fixed parameter", columns = vars(`(Intercept)`)) %>% 
    tab_style(style = cell_fill(color = "lightgrey"), locations = cells_body(columns = everything(),rows = delta <= 2)) %>% 
  tab_style(style = cell_fill(color = "darkgrey"), locations = cells_body( columns = everything(), rows = 1)) 

```


Specify best model:

```{r}
d_mod92la.bm<-lmer(LftotB_92~Sever +Lf_90+LfBtot_91+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)
```

Estimate marginal means/partial regression coefficients for the main effects:

```{r}
#Sever:
emmeans(d_mod92la.bm, pairwise~Sever, type="response")
#Partial regression coefficient for Lf_90 and LfBtot_91 can be found in the model summary table:
summary(d_mod92la.bm)
```


