---
title: "Plots"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

This document contains the code needed to recreate the plots in the manuscript (Figure 1, Figure 2, Figure S1). *Note*: All figure aesthetics were finalized in Adobe Illustrator.

```{r}
library(cowplot)
library(mdthemes)
```

## Figure 1

*Note*: the models used to generate the estimated marginal means were the best models returned by our model selection procedure. See `Model selection analyses.Rmd`.

### Panels A to C

Working with leaf area in 1990:
```{r, warning=FALSE, message=FALSE}
may<-read.csv("Datasets/mayapple_1990_updated.csv")
##need to make the variables we are working with numeric:
may$Lf_90<-as.numeric(as.character(may$Lf_90))

may<-may %>% 
  filter(!is.na(Lf_90))

model.bm<-lmer(Lf_90~Sex_90*Time+Sever*Time+(1|COLONY), data=may, na.action = "na.fail", REML=FALSE)
```

Drawing panels:

```{r, warning=FALSE, message=FALSE}

Sever_1 <- emmip(model.bm,  ~ Sever ,CIs = TRUE)+xlab("")+mdthemes::md_theme_classic() +ggtitle("**Sever**  
                                                                                                         *Leaf area in 1990*")
Time_1 <- emmip(model.bm,  ~ Time ,CIs = TRUE)+xlab("")+ylab("")+mdthemes::md_theme_classic() +ggtitle("**Time**") 

Int_1 <- emmip(model.bm, Time  ~ Sever  ,CIs = TRUE)+xlab("")+ylab("")+mdthemes::md_theme_classic()+ggtitle("**Sever x Time**")
```

### Panels D and E

Working with front shoot production in 1991:

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-readxl::read_excel("Datasets/mayapple_1990_updated.xlsx", na=".")

# select variables for analysis and filter out rows that don't have a leaf area recorded in 1990
mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,brnoF_91,Sex_90,Sever,Time,SexF1_91,SexF2_91,SexF3_91,LfFtot_91) %>% 
  filter(!is.na(Lf_90))

# create binary variable for plants that are dead/have no branches (0) or have 1 or more branches (1)
# append new variable to the dataframe
mayappleData<- mayappleData %>% 
  dplyr::mutate(deadAliveBinary=ifelse(brnoF_91>0,1,0))

sur_91f_bm<-glmmTMB(deadAliveBinary~ Sever+Time +  (1|COLONY), data = mayappleData,
                family="binomial")


```

Drawing panels:

```{r}
Sever_2 <- emmip(sur_91f_bm,   ~ Sever  , type="response",CIs = TRUE)+xlab("")+ylab("Predicted probability")+mdthemes::md_theme_classic()+ggtitle("*Front survival in 1991*")

Time_2 <- emmip(sur_91f_bm,   ~ Time  ,type="response",CIs = TRUE)+xlab("")+ylab("")+mdthemes::md_theme_classic()
```

### Panel F

Working with front leaf area in 1991, using systems with a non-zero leaf area in 1991.

```{r, warning=FALSE, message=FALSE}
mayappleData$LfFtot_91<-as.numeric(mayappleData$LfFtot_91)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfFtot_91)) %>% 
  filter(LfFtot_91>0)

mod91laf.bm<-lmer(LfFtot_91~Sever+Lf_90 +(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

```

Drawing panel:

```{r}
Sever_3 <- emmip(mod91laf.bm,   ~ Sever  ,CIs = TRUE)+xlab("")+mdthemes::md_theme_classic()+ggtitle("*Front leaf area in 1991*")

```

### Panels G and H

Working with back shoot production in 1991:

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-readxl::read_excel("Datasets/mayapple_1990_updated.xlsx", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,Sex_90,Sever,Time,brnoB_91,LfB1_91,SexB1_91,LfB2_91,SexB2_91,LfBtot_91,Lftot_91) %>% 
  filter(!is.na(Lf_90))

mayappleData<- mayappleData %>% 
  dplyr::mutate(backShootBranchBinary=ifelse(brnoB_91>0,1,0)) %>% 
  filter(!is.na(backShootBranchBinary))

sur_91b_bm<-glmmTMB(backShootBranchBinary~ Sever+Time + Lf_90 +(1|COLONY), data = mayappleData,
                family="binomial")
```

Drawing panels:

```{r}
Sever_4 <- emmip(sur_91b_bm,   ~ Sever  ,type="response",CIs = TRUE )+xlab("")+ylab("Predicted probability")+mdthemes::md_theme_classic()+ggtitle("*Back shoot production in 1991*")

Time_4 <- emmip(sur_91b_bm,   ~ Time  ,type="response",CIs = TRUE )+xlab("")+ylab("")+mdthemes::md_theme_classic()
```

### Panels I and J:

Working with leaf area of the back of the system in 1991, for those systems with non-zero leaf area in the back in 1991.

```{r}
mayappleData$LfBtot_91<-as.numeric(mayappleData$LfBtot_91)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfBtot_91)) %>% 
  filter(LfBtot_91>0)

b_mod91la.bm<-lmer(LfBtot_91~Sex_90+Sever+Time +Lf_90+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

```

Drawing panels:

```{r}
Sever_5 <- emmip(b_mod91la.bm,  ~ Sever  ,CIs = TRUE )+mdthemes::md_theme_classic()+ggtitle("*Back leaf area in 1991*")
Time_5 <- emmip(b_mod91la.bm,  ~ Time  ,CIs = TRUE )+ylab("")+mdthemes::md_theme_classic()
```


### Making the composite figure:

```{r}
ggdraw() +
  draw_plot(Sever_1, x=0, y=0.8, width=0.33, height=0.2)+
  draw_plot(Sever_2, x=0, y=0.6,width=0.33, height=0.2)+
  draw_plot(Sever_3, x=0, y=0.4,width=0.33, height=0.2)+
  draw_plot(Sever_4, x=0, y=0.2,width=0.33, height=0.2)+
  draw_plot(Sever_5, x=0, y=0.0,width=0.33, height=0.2)+
  draw_plot(Time_1, x=0.33, y=0.8, width=0.33, height=0.2)+
  draw_plot(Time_2, x=0.33, y=0.6,width=0.33, height=0.2)+
  draw_plot(Time_4, x=0.33, y=0.2,width=0.33, height=0.2)+
  draw_plot(Time_5, x=0.33, y=0.0,width=0.33, height=0.2)+
  draw_plot(Int_1, x=0.67, y=0.8, width=0.33, height=0.2)
```


## Figure 2

```{r}

mayappleData91 <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_90,Sex_90,Sever,Time, 
                LfFtot_91, LfBtot_91, Lftot_91,
                LftotF_92, LftotB_92, Lftot_92,
                brnoF_91, brnoB_91,
                brnoF_92, brnoB_92) %>% 
  mutate(Lftot_92=parse_double(Lftot_92), Lftot_91=parse_double(Lftot_91)) %>% 
  mutate(LftotB_92=parse_double(LftotB_92), LfBtot_91=parse_double(LfBtot_91) ) %>% 
  mutate(LftotF_92=parse_double(LftotF_92), LfFtot_91=parse_double(LfFtot_91) ) %>% 
  mutate(brnoF_92=parse_double(brnoF_92), brnoB_92=parse_double(brnoB_92) ) %>% 
  filter(!is.na(Lf_90)) %>% 
  filter(!is.na(Lftot_91)) %>% 
mutate(Status = case_when(LfFtot_91!=0 & LfBtot_91!=0 ~"Two",
                                             LfFtot_91==0 & LfBtot_91!=0 ~"One",
                                             LfFtot_91!=0 & LfBtot_91==0 ~"One",
                                             LfFtot_91==0 & LfBtot_91==0 ~"None")) %>% 
  mutate(Status = factor(Status, levels=c("None", "One", "Two"))) %>% 
 filter(Status != "None") %>% 
  mutate(Lf91s=ifelse(Status=="Two", Lftot_91/2, Lftot_91)) 

s91mod1<-lmer(Lftot_91~Status*Sever+(1|COLONY), data = mayappleData91)
a<-emmip(s91mod1, Sever~Status  ,CIs = TRUE)+ ylab("Total leaf area")+ 
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+xlab("")+theme(legend.position = "none")+theme(panel.border = element_blank(), axis.line = element_line())

counts<-mayappleData91 %>% group_by(Status, Sever) %>% 
  count(N=length(Sever))

s91mod2<-lmer(Lf91s~Status*Sever+(1|COLONY), data = mayappleData91)
b<-emmip(s91mod2, Sever~Status  ,CIs = TRUE)+ ylab("Average leaf area per system")+ 
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+scale_y_continuous(limits=c(-250, 1500))+ geom_text(data=counts, aes(x=Status, y=N, label=N, color=Sever))



ggdraw(b) +
  draw_plot(a, .35, .47, .6, .6, scale=0.75)
```

## Figure S1

```{r}
mayappleData92 <- mayappleData91 %>% 
  filter(!is.na(Lftot_92)) %>% 
  mutate(Status = case_when(LftotF_92!=0 & LftotB_92!=0 ~"Two",
                                             LftotF_92==0 & LftotB_92!=0 ~"One",
                                             LftotF_92!=0 & LftotB_92==0 ~"One",
                                             LftotF_92==0 & LftotB_92==0 ~"None")) %>% 
  mutate(Status = factor(Status, levels=c("None", "One", "Two"))) %>% 
  mayappleData92 <- mayappleData92 %>% filter(Status != "None") %>% 
  mutate(Lf92s=ifelse(Status=="Two", Lftot_92/2, Lftot_92)) 


s92mod1<-lmer(Lftot_92~Status*Sever+(1|COLONY), data = mayappleData92)
a<-emmip(s92mod1, Sever~Status  ,CIs = TRUE)+ ylab("Total leaf area")+ 
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+xlab("")+theme(legend.position = "none")+theme(panel.border = element_blank(), axis.line = element_line())

counts<-mayappleData92 %>% group_by(Status, Sever) %>% 
  count(N=length(Sever))

s92mod2<-lmer(Lf92s~Status*Sever+(1|COLONY), data = mayappleData92)
b<-emmip(s92mod2, Sever~Status  ,CIs = TRUE)+ ylab("Average leaf area per system")+ 
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+scale_y_continuous(limits=c(-250, 1500))+ geom_text(data=counts, aes(x=Status, y=N, label=N, color=Sever))

ggdraw(b) +
  draw_plot(a, .35, .47, .6, .6, scale=0.75)

```


