---
title: "Plots"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

This document contains the code needed to recreate the data plots in the manuscript (Figure 2, Figure 3, Figure S1). *Note*: All figure aesthetics were finalized in Adobe Illustrator.

```{r}
library(tidyverse)
library(cowplot)
library(mdthemes)
library(ggpubr)
library(lme4)
library(emmeans)
library(glmmTMB)
library(gridExtra)
```

## Figure 2

*Note*: the models used to generate the estimated marginal means were the best models returned by our model selection procedure. See `Model selection analyses.Rmd`.

### Panels A to C

Working with leaf area in year $t$:
```{r, warning=FALSE, message=FALSE}
may<-read_csv("mayapple.csv", na=".")
##need to make the variables we are working with numeric:
may$Lf_t<-as.numeric(as.character(may$Lf_t))

may<-may %>% 
  filter(!is.na(Lf_t))

model.bm<-lmer(Lf_t~Sex_t*Time+Sever*Time+(1|COLONY), data=may, na.action = "na.fail", REML=FALSE)
```

Drawing panels:

```{r, warning=FALSE, message=FALSE}

Sever_1 <- emmip(model.bm,  ~ Sever ,CIs = TRUE)+xlab("")+mdthemes::md_theme_classic() +ggtitle("**Sever**  
                                                                                                         *Leaf area in year t*") +
  ylab("Leaf area (cm^2)")+ xlab("Levels of Sever")

Time_1 <- emmip(model.bm,  ~ Time ,CIs = TRUE)+xlab("")+ylab("")+mdthemes::md_theme_classic() +ggtitle("**Time**") 

Int_1 <- emmip(model.bm, Time  ~ Sever  ,CIs = TRUE)+xlab("")+ylab("")+mdthemes::md_theme_classic()+ggtitle("**Sever x Time**")
```

### Panels D and E

Working with front shoot production in year $t+1$:

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

# select variables for analysis and filter out rows that don't have a leaf area recorded in year t
mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,brnoF_t1,Sex_t,Sever,Time,LfFtot_t1) %>% 
  filter(!is.na(Lf_t))

# create binary variable for plants that are dead/have no branches (0) or have 1 or more branches (1)
# append new variable to the dataframe
mayappleData<- mayappleData %>% 
  dplyr::mutate(deadAliveBinary=ifelse(brnoF_t1>0,1,0))

sur_t1f_bm<-glmmTMB(deadAliveBinary~ Sever+Time +  (1|COLONY), data = mayappleData,
                family="binomial")


```

Drawing panels:

```{r}
Sever_2 <- emmip(sur_t1f_bm,   ~ Sever  , type="response",CIs = TRUE)+xlab("")+ylab("Probability of shoot production")+mdthemes::md_theme_classic()+ggtitle("*Front survival in year t+1*")

Time_2 <- emmip(sur_t1f_bm,   ~ Time  ,type="response",CIs = TRUE)+xlab("")+ylab("")+mdthemes::md_theme_classic()
```

### Panel F

Working with front leaf area in year $t+1$, using systems with a non-zero leaf area in year $t+1$.

```{r, warning=FALSE, message=FALSE}
mayappleData$LfFtot_t1<-as.numeric(mayappleData$LfFtot_t1)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfFtot_t1)) %>% 
  filter(LfFtot_t1>0)

mod_t1laf.bm<-lmer(LfFtot_t1~Sever+Lf_t +(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

```

Drawing panel:

```{r}
Sever_3 <- emmip(mod_t1laf.bm,   ~ Sever  ,CIs = TRUE)+xlab("")+mdthemes::md_theme_classic()+ggtitle("*Front leaf area in year t+1*") + ylab("Leaf area (cm^2)")

```

### Panels G and H

Working with back shoot production in year $t+1$:

```{r}
# set na.strings="." to recode '.' to NA
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time,brnoB_t1,LfBtot_t1,Lftot_t1) %>% 
  filter(!is.na(Lf_t))

mayappleData<- mayappleData %>% 
  dplyr::mutate(backShootBranchBinary=ifelse(brnoB_t1>0,1,0)) %>% 
  filter(!is.na(backShootBranchBinary))

sur_t1b_bm<-glmmTMB(backShootBranchBinary~ Sever+Time + Lf_t +(1|COLONY), data = mayappleData,
                family="binomial")
```

Drawing panels:

```{r}
Sever_4 <- emmip(sur_t1b_bm,   ~ Sever  ,type="response",CIs = TRUE )+xlab("")+ylab("Probability of shoot production")+mdthemes::md_theme_classic()+ggtitle("*Back shoot production in year t+1*")

Time_4 <- emmip(sur_t1b_bm,   ~ Time  ,type="response",CIs = TRUE )+xlab("")+ylab("")+mdthemes::md_theme_classic()
```

### Panels I and J:

Working with leaf area of the back of the system in year $t+1$, for those systems with non-zero leaf area in the back in year $t+1$.

```{r}
mayappleData$LfBtot_t1<-as.numeric(mayappleData$LfBtot_t1)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfBtot_t1)) %>% 
  filter(LfBtot_t1>0)

b_mod_t1la.bm<-lmer(LfBtot_t1~Sex_t+Sever+Time +Lf_t+(1|COLONY), data = mayappleData, na.action ="na.fail", REML=FALSE)

```

Drawing panels:

```{r}
Sever_5 <- emmip(b_mod_t1la.bm,  ~ Sever  ,CIs = TRUE )+mdthemes::md_theme_classic()+ggtitle("*Back leaf area in year t+1*") + ylab("Leaf area (cm^2)") + xlab("Levels of Sever")
Time_5 <- emmip(b_mod_t1la.bm,  ~ Time  ,CIs = TRUE )+ylab("")+mdthemes::md_theme_classic() + xlab("Levels of Time")
```


### Making the composite figure:

```{r}
ggdraw() +
  draw_plot(Sever_1, x=0, y=0.8, width=0.33, height=0.2)+
  draw_plot(Sever_2, x=0, y=0.6,width=0.33, height=0.2)+
  draw_plot(Sever_3, x=0, y=0.4,width=0.33, height=0.2)+
  draw_plot(Sever_4, x=0, y=0.2,width=0.33, height=0.2)+
  draw_plot(Sever_5, x=0, y=0.0,width=0.33, height=0.2)+
  draw_plot(Time_1, x=0.33, y=0.8, width=0.33, height=0.2)+
  draw_plot(Time_2, x=0.33, y=0.6,width=0.33, height=0.2)+
  draw_plot(Time_4, x=0.33, y=0.2,width=0.33, height=0.2)+
  draw_plot(Time_5, x=0.33, y=0.0,width=0.33, height=0.2)+
  draw_plot(Int_1, x=0.67, y=0.8, width=0.33, height=0.2)
```


## Figure 3

```{r}
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")


mayappleData_t1 <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time, 
                LfFtot_t1, LfBtot_t1, Lftot_t1,
                LftotF_t2, LftotB_t2, Lftot_t2,
                brnoF_t1, brnoB_t1,
                brnoF_t2, brnoB_t2) %>% 
  mutate(Lftot_t2=parse_double(Lftot_t2), Lftot_t1=parse_double(Lftot_t1)) %>% 
  mutate(LftotB_t2=parse_double(LftotB_t2), LfBtot_t1=parse_double(LfBtot_t1) ) %>% 
  mutate(LftotF_t2=parse_double(LftotF_t2), LfFtot_t1=parse_double(LfFtot_t1) ) %>% 
  mutate(brnoF_t2=parse_double(brnoF_t2), brnoB_t2=parse_double(brnoB_t2) ) %>% 
  filter(!is.na(Lf_t)) %>% 
  filter(!is.na(Lftot_t1)) %>% 
mutate(Status = case_when(LfFtot_t1!=0 & LfBtot_t1!=0 ~"Two",
                                             LfFtot_t1==0 & LfBtot_t1!=0 ~"One",
                                             LfFtot_t1!=0 & LfBtot_t1==0 ~"One",
                                             LfFtot_t1==0 & LfBtot_t1==0 ~"None")) %>% 
  mutate(Status = factor(Status, levels=c("None", "One", "Two"))) %>% 
 filter(Status != "None") %>% 
  mutate(Lf_t1s=ifelse(Status=="Two", Lftot_t1/2, Lftot_t1)) 

s_t1mod1<-lmer(Lftot_t1~Status*Sever+(1|COLONY), data = mayappleData_t1)
a<-emmip(s_t1mod1, Sever~Status  ,CIs = TRUE,linearg = list(linetype = NA))+mdthemes::md_theme_classic()+ ylab("Total leaf area (cm^2)")+ 
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+xlab("")+theme(legend.position = "none")+theme(panel.border = element_blank(), axis.line = element_line())

counts<-mayappleData_t1 %>% group_by(Status, Sever) %>% 
  count(N=length(Sever))

s_t1mod2<-lmer(Lf_t1s~Status*Sever+(1|COLONY), data = mayappleData_t1)
b<-emmip(s_t1mod2, Sever~Status  ,CIs = TRUE, linearg = list(linetype = NA))+mdthemes::md_theme_classic()+ ylab("Average leaf area per system (cm^2)")+ xlab("Number of descendant rhizome systems") +
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+scale_y_continuous(limits=c(-250, 1500))+ geom_text(data=counts, aes(x=Status, y=N, label=N, color=Sever))+theme(panel.border = element_blank(), axis.line = element_line())

grid.arrange(a, b, nrow=1)

ggdraw() +
  draw_plot(a, x=0, y=0, width=0.5, height=1)+
  draw_plot(Sever_2, x=0.5, y=0,width=0.5, height=1)
```

## Figure S3

```{r}




#This is year t+1 front leaf area. We need to do this for all of the response variables discussed for Q3
mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,brnoF_t1,Sex_t,Sever,Time,LfFtot_t1) %>% 
  filter(!is.na(Lf_t))

mayappleData$LfFtot_t1<-as.numeric(mayappleData$LfFtot_t1)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfFtot_t1))

a <- ggplot(data=mayappleData, aes(x=Lf_t, y=LfFtot_t1, color=Sex_t)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)+theme_classic()+
  ylab("Front leaf area in year t+1") +xlab("Leaf area in year t")+ scale_color_discrete(name="Sexual status in year t")

# year t+1 back shoot production

mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time,brnoB_t1,LfBtot_t1,Lftot_t1) %>% 
  filter(!is.na(Lf_t))

mayappleData<- mayappleData %>% 
  dplyr::mutate(backShootBranchBinary=ifelse(brnoB_t1>0,1,0)) %>% 
  filter(!is.na(backShootBranchBinary))

b <- ggplot(data=mayappleData, aes(x=Lf_t, y=backShootBranchBinary, color=Sex_t)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)+theme_classic()+
  ylab("Probability of producing \n back shoot in year t+1")  +xlab("Leaf area in year t")+ scale_color_discrete(name="Sexual status in year t")


# year t+1 back leaf area

mayappleData$LfBtot_t1<-as.numeric(mayappleData$LfBtot_t1)

mayappleData <- mayappleData %>% 
  filter(!is.na(LfBtot_t1)) %>% 
  filter(LfBtot_t1>0)

c <- ggplot(data=mayappleData, aes(x=Lf_t, y=LfBtot_t1, color=Sex_t)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)+theme_classic()+
  ylab("Back leaf area in year t+1")  +xlab("Leaf area in year t")+ scale_color_discrete(name="Sexual status in year t")


# year t+2 front leaf area

# need leaf year t and front leaf year t+1

mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time,LftotF_t2, Lftot_t1, LfBtot_t1, LfFtot_t1) %>% 
  mutate(LftotF_t2=parse_double(LftotF_t2), Lftot_t1=parse_double(Lftot_t1), LfBtot_t1=parse_double(LfBtot_t1), LfFtot_t1=parse_double(LfFtot_t1) ) %>% 
  filter(!is.na(Lf_t)) %>% 
  filter(!is.na(Lftot_t1)) %>% 
  filter(!is.na(LftotF_t2))

d <- ggplot(data=mayappleData, aes(x=Lf_t, y=LftotF_t2, color=Sex_t)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)+theme_classic()+
  ylab("Front leaf area in year t+2") +xlab("Leaf area in year t")+ scale_color_discrete(name="Sexual status in year t")



# year t+2 back leaf area

mayappleDatasetFull<-read_csv("mayapple.csv", na=".")

mayappleData <- mayappleDatasetFull %>% 
  dplyr::select(COLONY,Lf_t,Sex_t,Sever,Time,LftotB_t2, Lftot_t1, LfBtot_t1, LfFtot_t1) %>% 
  mutate(LftotB_t2=parse_double(LftotB_t2), Lftot_t1=parse_double(Lftot_t1), LfBtot_t1=parse_double(LfBtot_t1), LfFtot_t1=parse_double(LfFtot_t1) ) %>% 
  filter(!is.na(Lf_t)) %>% 
  filter(!is.na(Lftot_t1)) %>% 
  filter(!is.na(LftotB_t2))

e <- ggplot(data=mayappleData, aes(x=Lf_t, y=LftotB_t2, color=Sex_t)) +
  geom_point() + geom_smooth(method = "lm", se = FALSE)+theme_classic()+
  ylab("Back leaf area in year t+2") +xlab("Leaf area in year t")+ scale_color_discrete(name="Sexual status in year t")



ggarrange(a,b,c,d,e, nrow=3, ncol=2, common.legend = TRUE, legend = "bottom", labels="AUTO", label.x=0.15, hjust=0)

ggdraw() +
  draw_plot(a, x=0, y=0.75, width=0.5, height=0.25)+
  draw_plot(c, x=0, y=0.5, width=0.5, height=0.25)+
  draw_plot(e, x=0, y=0.25, width=0.5, height=0.25)+
  draw_plot(g, x=0, y=0, width=0.5, height=0.25)+
  draw_plot(b, x=0.5, y=0.75, width=0.5, height=0.25)+
  draw_plot(d, x=0.5, y=0.5, width=0.5, height=0.25)+
  draw_plot(f, x=0.5, y=0.25, width=0.5, height=0.25)
```



## Figure S2

```{r}
mayappleData_t2 <- mayappleData_t1 %>% 
  filter(!is.na(Lftot_t2)) %>% 
  mutate(Status = case_when(LftotF_t2!=0 & LftotB_t2!=0 ~"Two",
                                             LftotF_t2==0 & LftotB_t2!=0 ~"One",
                                             LftotF_t2!=0 & LftotB_t2==0 ~"One",
                                             LftotF_t2==0 & LftotB_t2==0 ~"None")) %>% 
  mutate(Status = factor(Status, levels=c("None", "One", "Two")))

  mayappleData_t2 <- mayappleData_t2 %>% filter(Status != "None") %>% 
  mutate(Lf_t2s=ifelse(Status=="Two", Lftot_t2/2, Lftot_t2)) 


s_t2mod1<-lmer(Lftot_t2~Status*Sever+(1|COLONY), data = mayappleData_t2)
a<-emmip(s_t2mod1, Sever~Status  ,CIs = TRUE,linearg = list(linetype = NA))+ ylab("Total leaf area (cm^2)")+ 
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+xlab("")+theme(legend.position = "none")+theme(panel.border = element_blank(), axis.line = element_line())

counts<-mayappleData_t2 %>% group_by(Status, Sever) %>% 
  count(N=length(Sever))

s_t2mod2<-lmer(Lf_t2s~Status*Sever+(1|COLONY), data = mayappleData_t2)
b<-emmip(s_t2mod2, Sever~Status  ,CIs = TRUE,linearg = list(linetype = NA))+ ylab("Average leaf area per system (cm^2)")+ xlab("Number of descendant rhizome systems") + 
  theme_bw()+theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank())+scale_y_continuous(limits=c(-250, 1500))+ geom_text(data=counts, aes(x=Status, y=N, label=N, color=Sever))

ggdraw(b) +
  draw_plot(a, .35, .47, .6, .6, scale=0.75)

```

## Figure S1


```{r load year_t, warning=FALSE, message=FALSE}
may<-read_csv("mayapple.csv",na=".")

may$Lf_t<-as.numeric(as.character(may$Lf_t))

may<-may %>% 
  filter(!is.na(Lf_t))

model.bm<-lmer(Lf_t~Sex_t*Time+Sever*Time+(1|COLONY), data=may, na.action = "na.fail", REML=FALSE)
```

```{r}
#Drawing the figure

sex_sever <- emmip(model.bm,  Time~Sex_t,CIs = TRUE, linearg = list(linetype = NA))+xlab("")+mdthemes::md_theme_classic() +ggtitle("**Sex  x severing time**") + xlab("Sexual status") + ylab("Leaf area ( cm^2 )") 
                                                                                            
```

